\cursedforest\ - A Random Forest Implementation for ``Big'' and ``Wide'' Data \\ Supplementary Information

### 2017-04-26 doc/publications/plos16/nm.tex
small changes to nm.tex

Analyses are being performed on ever larger datasets with unprecedentedly detailed information per data point. Specifically in
genomics, cohort sizes are in the thousands, each comprising of millions of variants that make up a person's genetic
differences. Traditional approaches are unable to cater for this task and even standard implementations using novel
compute paradigms, such as Hadoop/Spark, are limited by the finite resource of computer memory on such
data. Specifically, Spark ML's supervised machine learning methods suffer from ``the curse of dimensionality''. That is,
they scale well with samples ($n$) but not with features ($p$, variants).

We introduce \cursedforest, a tailored implementation of random forests, designed to handle data with extremely large
number of variables per sample. We showcase \cursedforest to accurately predict ethnicity from whole genome sequencing
profiles of 2,500 individuals and 80 million genomic variants each (1000 Genomes Project) in 30 minutes. \cursedforest
was the only implementation scaling to the full dataset, outperforming other methods on smaller subsets by 80\% or
more. We also demonstrate a GWAS-style analysis using \cursedforest's feature selection approach on an imputed SNP array
dataset (2000 x 7 Million) on Bone mineral density (BMD), successfully predicting known BMD-genes as well as identifying
novel candidates.

@@

\section{Introduction}
The digital revolution is seeing a dramatic increase in data collected about almost every aspect of
life~\cite{Loebbecke2015}.  These datasets are not only growing vertically, by capturing more events, but also
horizontally by capturing more information about these events.  The challenge of big and ``wide'' data is especially
pronounced in the health space where, for example, whole genome sequencing (WGS) technology enables researchers to
interrogate all 3 billion base pairs of the human genome.

@@

Hence using more sophisticated machine learning (ML) approaches, in particular tree-based models, have been successful
for taking the interaction of variables into account~\cite{Wright.et.al.2016}. In addition, random forests are well
suited for processing ``wide'' genomic data for two reasons.  Firstly, while other machine learning applications have
the propensity to overfit datasets with more features $p$ than samples $n$ (a consequence of the ``curse of
dimensionality''~\cite{Bauer2014, bellman1961adaptive}), decision trees are resistant to overfitting.  Secondly, random
forests are also very easy to parallelise. As the forest is a sum of decision trees, it is possible to grow separate
trees on different processors and combine the trees. 
%See the discussion of the properties of random forests in Section \ref{section:methods}.

The most popular random forest software packages to date are R-based, although it has been reported that current implementations 
do not scale well with increasing feature size~\cite{Wright.and.Ziegle.2016}.  This is despite R now supporting long vectors (length greater
than $2^{31}-1$), and the \randomforest implementation in R lending itself to parallelisation by growing multiple
trees simultaneously and combining them for the final result~\cite{Liaw.and.Weiner.2002}.  More suitable implementations for ``wide''
genomic data have been developed in other programming languages. \textsc{CloudForest} \cite{Bressler2015}, written in Go,
achieves fast run times by effective use of the CPU cache, optimising for different classes of features and efficiently
multithreading.  Similarly, \ranger~\cite{Wright.and.Ziegle.2016}, written in C++ with an R front-end, is
specifically optimised for dealing with large datasets.

@@

Despite overcoming the node-boundary limitation, the standard implementation of random forest in Spark ML
is not able to handle the extremely ``wide'' genomic data as it was developed for a large number of samples
with only modest dimensionality~\cite{NIPS2016_6366}.  Although Spark ML can build a random forest model on a subset of the data (chromosome
1), we show that the time taken is excessive due to the large amount of data being aggregated and processed by the driver node during
intermediate stages of building the model.  This unbalanced work load where the driver
node becomes the bottleneck and worker nodes being idle prevents a seamless scaling to larger datasets. We also show that the memory
requirements per executor increases with dimensionality due to the data types Spark ML uses.%, which we elaborate on in the discussion.

Here we introduce \cursedforest, a tailored Hadoop/Spark-based implementation of random forests specifically designed to cater
for ``big'' (many samples) and ``wide'' (many features) datasets. 
\cursedforest extends our previously developed variant interpretation framework, VariantSpark~\cite{OBrien2015}, to now offer
supervised as well as unsupervised ML algorithms in the Spark framework.
In our implementation a Spark application runs on a ``driver'' node and
distributes tasks to many ``worker'' nodes, or ``executors''. 
By also utilising Spark to read in and manipulate the standard genomic variant format (VCF)
directly, \cursedforest outperforms existing tools even on small datasets where multithreading generally performs
well. Harnessing the virtually unlimited capability to parallelise tasks, \cursedforest can hence explore the solution 
space faster by building a larger number of diverse models to generate a consensus classifier.

Using this facility, \cursedforest is capable of parallelising the split for each node in a tree thereby handling millions 
of features, as required to process whole genome sequencing data or SNP array data with unobserved genotypes 
imputed~\cite{Howie2012}.  This provides the potential to generate datasets of hundreds of thousands of individuals 
with millions of variants (imputing the GWAS catalog), highlighting the need for modern
compute paradigms in the genomics space.

@@

Random forests apply the technique of bootstrap aggregation or bagging to decision trees.  The training datasets are
independently drawn, with replacement, from the joint distribution of $(X,Y)$ so that each set comprises $n$
$(p+1)$-tuples $(x_1,y_1),\ldots, (x_n,y_n)$. This is done $B$ times and a tree is grown on each of these $B$ samples.
The results are then combined in an appropriate way. In a classification problem we take the majority vote for the trees
over the $q=1,..,Q$ classes,
\begin{equation*}
{{h_f}}= \arg \max_q \left(\sum_b I(h(x;B_b)=q)\right).
\end{equation*}
In addition to growing the trees
on bootstrapped samples, each node in the tree is split on a randomly
selected subset of the variables.

@@

\cite{Strobl.et.al.2007} introduce changes to the random forest algorithm to fix 1) and 2).  Effect 1) will be apparent
in the \cursedforest algorithm as well.  However, there are many examples, particularly in genomics, where we have data
sets that are:
\begin{itemize}
\item very wide;
\item all variables have the same number of levels (for example, measurements at different points along the genome);
\item the hypothesized functional relationship is of a form amenable to a decision tree.
\end{itemize}
GWAS datasets are an obvious application area for this algorithm.

@@

However, within an RDD, Spark ML stores each sample as a vector. Unlike RDDs, which can be partitioned and distributed
across multiple nodes, each vector must be present in its entirety on any node accessing it. This is no problem with
typical datasets; however as dimensionality increases, the vectors eventually reach a size where they can no longer fit
into a single node's memory.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This section talks about the datasets
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Datasets}
\subsubsection{1000 Genomes Project}
We obtained the 1000 Genomes Project data as VCF files from their FTP site.  Each VCF file contains the variants for
every individual across one chromosome.  We use the phase 3 dataset which contains 2,504 individuals with approximately
81 million variant positions (i.e. a matrix of $2,504 \times 81$ million).  For compatibility with the majority of the
algorithms we demonstrate, we convert the original variant representations to double representations.  For example, we
store the absence of a variant (0\textbar0) as 0, a heterozygous
variant (1\textbar0 or 0\textbar1) as 1, and a
homozygous variant (1\textbar1) as 2, so we have an ``additive'' model.  However, this is not the only possible encoding \cite{Goldstein.et.al.2011}. Note that from
the encodings listed in Table~\ref{table:encodings}, the additive encoding will present the most difficulty for a random
forest model.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 2: Feature Selection on real data 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection{GWAS-style analysis using \cursedforest identifies novel variants associated with Bone Mineral Density}
\subsection{GWAS-style analyses using \cursedforest faithfully identify known disease-accociated genes}
In this section we use \cursedforest to perform feature selection on over 7.2 million genomic variants and identify the
genomic locations associated with Bone Mineral Density (BMD)~\cite{Duncan.2011}. 
%We build a model on the BMD dataset as described in Methods using a binary high or low BMD value as the label.  
We rank variants by their Gini impurity, with the most important one ranked at position 1, and return the top 100,000 variants.

@@

\begin{figure}[tbhp]
  \begin{adjustwidth}{-2.00in}{0in}
    \caption{\textbf{Genome-wide importance of SNPs for Bone Mineral Density association}}
    \label{figure:ranksumtest}
     \begin{subfigure}[b]{0.5\linewidth}
     \includegraphics[totalheight=7cm]{./figs/BMDTopWholeGenome_annotated.png}
      \label{figure:BMDTopWholeGenome} 
    \end{subfigure} 
 \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[totalheight=7cm]{./figs/BMDTopWholeGenome_BW_random.png}
      \label{zoomplotTNFRSF} 
   \end{subfigure} 
    \begin{subfigure}[b]{0.5\linewidth}
    \includegraphics[totalheight=7cm]{./figs/new_10k_TNFRSF11B.pdf}
%     \includegraphics[totalheight=7cm]{./figs/new_10k_PRODH.pdf}
      \label{zoomplotprodh} 
    \end{subfigure} 
        \begin{subfigure}[b]{0.5\linewidth}
     \includegraphics[totalheight=7cm]{./figs/EUR_chr7_156_2-156_6mb.pdf}
      \label{zoomplotlinc} 
    \end{subfigure} 
    \begin{flushleft}
      Genome-wide SNPs and their importance for the BMD dataset {\bf(a)}; highlighted are the 100 literature assigned BMD genes (green) and four novel loci. The horizontal line indicates the top 99.98 FDR. 
      {\bf(b)} shows a previous GWAS analysis using logistic regression. 
      Locus Zoomplot for {\bf(c)} the top ranked gene TNFRSF11B and the two putative novel associations {\bf(d)} LINC01006. Each point represents a variant with the normalised Gini impurity score on the left y axis and LD to the central SNP in colour with the highest $r^2$ in red.
     \end{flushleft}
  \end{adjustwidth}
        \label{manhattenzoom} 
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 3: Classification selection
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest outperforms existing methods for classification problems}
In this section we demonstrate \cursedforest's classification accuracy on biological data and compare its performance against other published methods. 
We do this by predicting the ethnicity of the individuals in the 1000 Genomes Project from their genomic profiles and evaluate against the annotated label. 
We record the prediction error (OOB) as well as the runtime for each method. 
%@Aidan is 12 correct?
We limit the number of executors to 12 that are available to \cursedforest to keep the comparison fair for all methods.

As shown in Fig.~\ref{figure:1000genomes}, only \cursedforest is able to scale to the full phase3\_chr1-22 dataset (2,504 samples x 19,328,051 variants), with
the standard random forest implementation in R running out of resources on the relatively low-dimensional
phase1\_chr22 dataset (1,092 x 490,036). 
Ranger's c++ implementation and the R front-end version of it are able to process the second largest dataset (phase3\_chr1-3 with 2,504 x 19,328,051), however, both are unable to process larger datasets successfully. 
While the c++ implementation runs out of memory the R front-end, which also covers the loading and pre-processing, needs an unreasonable compute time for this task (>20h), see Supplementary Table 1.
The standard implementation of random forest in SparkML already runs out of memory after processing the third largest dataset, phase3\_chr1 (2,504 x 6,450,364). 
This demonstrates that Spark ML, although designed to handle many samples, is unable to efficiently cope with large numbers of features.

All tested methods have a worse runtime than \cursedforest for the datasets they processed successfully, except for the smallest dataset where the overhead of Spark's distributed platform made ranger's R implementation faster with 1 min 10 seconds compared to 1 min 50 seconds. For example, on phase3\_chr1-3 \cursedforest finishes in 1 hour, with ranger(R) requiring 4 hours and SparkML already needing 8 hours on a smaller dataset. 
Extrapolating from this, all comparable implementations hence would require more than 24 hours to finish, assuming the use of larger memory nodes.

@@

This may be because a large \mtry leads to more correlated trees as the same important features have a higher chance of
being selected in all, not yielding good performance outcomes.  This issue is less pronounced for classification error
where random features can mimic the response variable, hence resulting more in a performance plateau.  Increasing the
number of trees on the other hand improves performance especially when the trees are kept diverse (small \mtry) but
appropriate for large feature datasets (\mtry larger than default). \cursedforest's feature of parallelizing trees at
the node level hence caters perfectly to the requirement of large feature datasets as more trees can be built given a
fixed time budget.

@@

In order to asses these characteristics, we run \cursedforest classification on synthetic datasets with varying numbers
of variables (features) and samples, similar to the dataset used in \cite{Wright.and.Ziegle.2016} to evaluate
\ranger, allocating varying number of CPU cores to the \cursedforest and also varying the computational complexity
of the random forests by using a range of \mtry values.

We investigate the different synthetic datasets generated for section \ref{section:synthetic_data} and measured the time
taken to build a random forest model of 100 trees. The results reported below are averages of 5 runs, and all the cases
were executed with the same random seed, to improve the consistency of measurements.

@@

To conclude \cursedforest is capable of processing 60 trees per hour on a datasets with 50 million variables and 10,000
samples, which is the size range for whole genome sequencing experiments of clinically relevant cohort sizes.

@@

\begin{figure}[tbhp]
  \begin{adjustwidth}{-1.00in}{0in}
    \caption{\textbf{Scalablity of the Wide Random Forest on synthetic datasets with varying number of samples and variables.}}
    \label{figure:synth2}
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[totalheight=6cm]{./figs/nvars_nsamples.png} 
      \caption{Number of trees build per hour when growing number of variables.} 
      \label{figure:synth2.a} 
      \vspace{4ex}
    \end{subfigure} 
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[totalheight=6cm]{./figs/nsamples_nvars.png}
      \caption{Number of trees build per hour when growing number of samples.}
    \label{figure:synth2.b}
    \vspace{4ex}
  \end{subfigure}
\end{adjustwidth}
\end{figure}

@@

Using a novel parallelization approach enabled by Spark, we extended random forests to cope with extremely high dimensional data (large number of variables). 
We have demonstrated that for datasets with more variable ($p$) than samples ($n$), such as GWAS applications, \cursedforest can perform effective variable selection replicating previous findings and uncovering novel associations in a Bone Mineral Density study. 
We are able to elevate the signal compared to previous approaches using logistic regression due to random forests also modelling interacting features.

@@

To showcase the breadth of application scope, we also demonstrate that \cursedforest's implementation aids in classification problems such as predicting the ethnicity of the 1000 genome project. 
By comparing \cursedforest to other implementations (including those optimised for large datasets) we have demonstrated that runtime can be improved by 80\% allowing analysis in timeframes relevant for point-of-care diagnostics such as for "Patients like mine" scenarios thereby enabling evidence based self learning health care.

@@

% We have demonstrated that using a different parallelization model can extend random forests to the case of an
% extremely large number of variables. We have treated the case of variable selection in a $p >> n$ model where most of
% the variables are uninformative and have demonstrated the utility of the model for large GWAS datasets. By comparing
% this implementation to other implementations (including those optimized for large datasets) we have demonstrated the
% utility of this approach.

>>>>

Analyses are being performed on ever larger data sets with unprecedentedly detailed information per data point. Specifically in
genomics, cohort sizes are in the thousands, each comprising of millions of variants that make up a person's genetic
differences. Traditional approaches are unable to cater for this task and even standard implementations using novel
compute paradigms, such as Hadoop/Spark, are limited by the finite resource of computer memory on such
data. Specifically, Spark ML's supervised machine learning methods suffer from ``the curse of dimensionality''. That is,
they scale well with samples ($n$) but not with features ($p$, variants).

We introduce \cursedforest, a tailored implementation of random forests, designed to handle data with extremely large
number of variables per sample. We showcase \cursedforest to accurately predict ethnicity from whole genome sequencing
profiles of 2,500 individuals and 80 million genomic variants each (1000 Genomes Project) in 30 minutes. \cursedforest
was the only implementation scaling to the full dataset, outperforming other methods on smaller subsets by 80\% or
more. We also demonstrate a GWAS-style analysis using \cursedforest's feature selection approach on an imputed SNP array
dataset (2000 x 7 Million) on bone mineral density (BMD), successfully predicting known BMD-genes as well as identifying
novel candidates.

@@

\section{Introduction}
The digital revolution is seeing a dramatic increase in data collected about almost every aspect of
life~\cite{Loebbecke2015}.  These data sets are not only growing vertically, by capturing more events, but also
horizontally by capturing more information about these events.  The challenge of big and ``wide'' data is especially
pronounced in the health space where, for example, whole genome sequencing (WGS) technology enables researchers to
interrogate all 3 billion base pairs of the human genome.

@@

Hence using more sophisticated machine learning (ML) approaches, in particular tree-based models, have been successful
for taking the interaction of variables into account~\cite{Wright.et.al.2016}. In addition, random forests are well
suited for processing ``wide'' genomic data for two reasons.  Firstly, while other machine learning applications have
the propensity to overfit data sets with more features $p$ than samples $n$ (a consequence of the ``curse of
dimensionality''~\cite{Bauer2014, bellman1961adaptive}), decision trees are resistant to overfitting.  Secondly, random
forests are also very easy to parallelise. As the forest is a sum of decision trees, it is possible to grow separate
trees on different processors and combine the trees. 
%See the discussion of the properties of random forests in Section \ref{section:methods}.

The most popular random forest software packages to date are R-based, although it has been reported that current implementations 
do not scale well with increasing feature size~\cite{Wright.and.Ziegle.2016}.  This is despite R now supporting long vectors (length greater
than $2^{31}-1$), and the \randomforest implementation in R lending itself to parallelisation by growing multiple
trees simultaneously and combining them for the final result~\cite{Liaw.and.Weiner.2002}.  More suitable implementations for ``wide''
genomic data have been developed in other programming languages. \textsc{CloudForest} \cite{Bressler2015}, written in Go,
achieves fast run times by effective use of the CPU cache, optimising for different classes of features and efficiently
multithreading.  Similarly, \ranger~\cite{Wright.and.Ziegle.2016}, written in C++ with an R front-end, is
specifically optimised for dealing with large data sets.

@@

Despite overcoming the node-boundary limitation, the standard implementation of random forest in Spark ML
is not able to handle the extremely ``wide'' genomic data as it was developed for a large number of samples
with only modest dimensionality~\cite{NIPS2016_6366}.  Although Spark ML can build a random forest model on a subset of the data (chromosome
1), we show that the time taken is excessive due to the large amount of data being aggregated and processed by the driver node during
intermediate stages of building the model.  This unbalanced work load where the driver
node becomes the bottleneck and worker nodes being idle prevents a seamless scaling to larger data sets. We also show that the memory
requirements per executor increases with dimensionality due to the data types Spark ML uses.%, which we elaborate on in the discussion.

Here we introduce \cursedforest, a tailored Hadoop/Spark-based implementation of random forests specifically designed to cater
for ``big'' (many samples) and ``wide'' (many features) data sets. 
\cursedforest extends our previously developed variant interpretation framework, VariantSpark~\cite{OBrien2015}, to now offer
supervised as well as unsupervised ML algorithms in the Spark framework.
In our implementation a Spark application runs on a ``driver'' node and
distributes tasks to many ``worker'' nodes, or ``executors''. 
By also utilising Spark to read in and manipulate the standard genomic variant format (VCF)
directly, \cursedforest outperforms existing tools even on small data sets where multithreading generally performs
well. Harnessing the virtually unlimited capability to parallelise tasks, \cursedforest can hence explore the solution 
space faster by building a larger number of diverse models to generate a consensus classifier.

Using this facility, \cursedforest is capable of parallelising the split for each node in a tree thereby handling millions 
of features, as required to process whole genome sequencing data or SNP array data with unobserved genotypes 
imputed~\cite{Howie2012}.  This provides the potential to generate data sets of hundreds of thousands of individuals 
with millions of variants (imputing the GWAS catalog), highlighting the need for modern
compute paradigms in the genomics space.

@@

Random forests apply the technique of bootstrap aggregation or bagging to decision trees.  The training data sets are
independently drawn, with replacement, from the joint distribution of $(X,Y)$ so that each set comprises $n$
$(p+1)$-tuples $(x_1,y_1),\ldots, (x_n,y_n)$. This is done $B$ times and a tree is grown on each of these $B$ samples.
The results are then combined in an appropriate way. In a classification problem we take the majority vote for the trees
over the $q=1,..,Q$ classes,
\begin{equation*}
{{h_f}}= \arg \max_q \left(\sum_b I(h(x;B_b)=q)\right).
\end{equation*}
In addition to growing the trees
on bootstrapped samples, each node in the tree is split on a randomly
selected subset of the variables.

@@

\cite{Strobl.et.al.2007} introduce changes to the random forest algorithm to fix 1) and 2).  Effect 1) will be apparent
in the \cursedforest algorithm as well.  However, there are many examples, particularly in genomics, where we have data
sets that are:
\begin{itemize}
\item very wide;
\item all variables have the same number of levels (for example, measurements at different points along the genome);
\item the hypothesized functional relationship is of a form amenable to a decision tree.
\end{itemize}
GWAS data sets are an obvious application area for this algorithm.

@@

However, within an RDD, Spark ML stores each sample as a vector. Unlike RDDs, which can be partitioned and distributed
across multiple nodes, each vector must be present in its entirety on any node accessing it. This is no problem with
typical data sets; however as dimensionality increases, the vectors eventually reach a size where they can no longer fit
into a single node's memory.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This section talks about the data sets
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Data Sets}
\subsubsection{1000 Genomes Project}
We obtained the 1000 Genomes Project data as VCF files from their FTP site.  Each VCF file contains the variants for
every individual across one chromosome.  We use the phase 3 dataset which contains 2,504 individuals with approximately
81 million variant positions (i.e. a matrix of $2,504 \times 81$ million).  For compatibility with the majority of the
algorithms we demonstrate, we convert the original variant representations to double representations.  For example, we
store the absence of a variant (0\textbar0) as 0, a heterozygous
variant (1\textbar0 or 0\textbar1) as 1, and a
homozygous variant (1\textbar1) as 2, so we have an ``additive'' model.  However, this is not the only possible encoding \cite{Goldstein.et.al.2011}. Note that from
the encodings listed in Table~\ref{table:encodings}, the additive encoding will present the most difficulty for a random
forest model.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 2: Feature Selection on real data 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection{GWAS-style analysis using \cursedforest identifies novel variants associated with Bone Mineral Density}
\subsection{GWAS-style analyses using \cursedforest faithfully identify known disease-accociated genes}
In this section we use \cursedforest to perform feature selection on over 7.2 million genomic variants and identify the
genomic locations associated with bone mineral density (BMD)~\cite{Duncan.2011}. 
%We build a model on the BMD dataset as described in Methods using a binary high or low BMD value as the label.  
We rank variants by their Gini impurity, with the most important one ranked at position 1, and return the top 100,000 variants.

@@

\begin{figure}[tbhp]
  \begin{adjustwidth}{-2.00in}{0in}
    \caption{\textbf{Genome-wide importance of SNPs for bone mineral density association}}
    \label{figure:ranksumtest}
     \begin{subfigure}[b]{0.5\linewidth}
     \includegraphics[totalheight=7cm]{./figs/BMDTopWholeGenome_annotated.png}
      \label{figure:BMDTopWholeGenome} 
    \end{subfigure} 
 \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[totalheight=7cm]{./figs/BMDTopWholeGenome_BW_random.png}
      \label{zoomplotTNFRSF} 
   \end{subfigure} 
    \begin{subfigure}[b]{0.5\linewidth}
    \includegraphics[totalheight=7cm]{./figs/new_10k_TNFRSF11B.pdf}
%     \includegraphics[totalheight=7cm]{./figs/new_10k_PRODH.pdf}
      \label{zoomplotprodh} 
    \end{subfigure} 
        \begin{subfigure}[b]{0.5\linewidth}
     \includegraphics[totalheight=7cm]{./figs/EUR_chr7_156_2-156_6mb.pdf}
      \label{zoomplotlinc} 
    \end{subfigure} 
    \begin{flushleft}
      Genome-wide SNPs and their importance for the BMD dataset {\bf(a)}; highlighted are the 100 literature assigned BMD genes (green) and four novel loci. The horizontal line indicates the top 99.98 FDR. 
      {\bf(b)} shows a previous GWAS analysis using logistic regression. 
      Locus Zoomplot for {\bf(c)} the top ranked gene TNFRSF11B and the two putative novel associations {\bf(d)} LINC01006. Each point represents a variant with the normalised Gini impurity score on the left y axis and LD to the central SNP in colour with the highest $r^2$ in red.
     \end{flushleft}
  \end{adjustwidth}
        \label{manhattenzoom} 
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 3: Classification selection
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest outperforms existing methods for classification problems}
In this section we demonstrate \cursedforest's classification accuracy on biological data and compare its performance against other published methods. 
We do this by predicting the ethnicity of the individuals in the 1000 Genomes Project from their genomic profiles and evaluate against the annotated label. 
We record the prediction error (OOB) as well as the runtime for each method. 
%@Aidan is 12 correct?
We limit the number of executors available to \cursedforest to 12 to keep the comparison fair for all methods.

As shown in Fig.~\ref{figure:1000genomes}, only \cursedforest is able to scale to the full phase3\_chr1-22 dataset (2,504 samples x 19,328,051 variants), with
the standard random forest implementation in R running out of resources on the relatively low-dimensional
phase1\_chr22 dataset (1,092 x 490,036). 
Ranger's c++ implementation and the R front-end version of it are able to process the second largest dataset (phase3\_chr1-3 with 2,504 x 19,328,051), however, both are unable to process larger data sets successfully. 
While the c++ implementation runs out of memory the R front-end, which also covers the loading and pre-processing, needs an unreasonable compute time for this task (>20h), see Supplementary Table 1.
The standard implementation of random forest in SparkML already runs out of memory after processing the third largest dataset, phase3\_chr1 (2,504 x 6,450,364). 
This demonstrates that Spark ML, although designed to handle many samples, is unable to efficiently cope with large numbers of features.

All tested methods have a worse runtime than \cursedforest for the data sets they processed successfully, except for the smallest dataset where the overhead of Spark's distributed platform made ranger's R implementation faster with 1 min 10 seconds compared to 1 min 50 seconds. For example, on phase3\_chr1-3 \cursedforest finishes in 1 hour, with ranger(R) requiring 4 hours and SparkML already needing 8 hours on a smaller dataset. 
Extrapolating from this, all comparable implementations hence would require more than 24 hours to finish, assuming the use of larger memory nodes.

@@

This may be because a large \mtry leads to more correlated trees as the same important features have a higher chance of
being selected in all, not yielding good performance outcomes.  This issue is less pronounced for classification error
where random features can mimic the response variable, hence resulting more in a performance plateau.  Increasing the
number of trees on the other hand improves performance especially when the trees are kept diverse (small \mtry) but
appropriate for large feature data sets (\mtry larger than default). \cursedforest's feature of parallelizing trees at
the node level hence caters perfectly to the requirement of large feature data sets as more trees can be built given a
fixed time budget.

@@

In order to asses these characteristics, we run \cursedforest classification on synthetic data sets with varying numbers
of variables (features) and samples, similar to the dataset used in \cite{Wright.and.Ziegle.2016} to evaluate
\ranger, allocating varying number of CPU cores to the \cursedforest and also varying the computational complexity
of the random forests by using a range of \mtry values.

We investigate the different synthetic data sets generated for section \ref{section:synthetic_data} and measured the time
taken to build a random forest model of 100 trees. The results reported below are averages of 5 runs, and all the cases
were executed with the same random seed, to improve the consistency of measurements.

@@

To conclude \cursedforest is capable of processing 60 trees per hour on a data sets with 50 million variables and 10,000
samples, which is the size range for whole genome sequencing experiments of clinically relevant cohort sizes.

@@

\begin{figure}[tbhp]
  \begin{adjustwidth}{-1.00in}{0in}
    \caption{\textbf{Scalablity of the Wide Random Forest on synthetic data sets with varying number of samples and variables.}}
    \label{figure:synth2}
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[totalheight=6cm]{./figs/nvars_nsamples.png} 
      \caption{Number of trees build per hour when growing number of variables.} 
      \label{figure:synth2.a} 
      \vspace{4ex}
    \end{subfigure} 
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[totalheight=6cm]{./figs/nsamples_nvars.png}
      \caption{Number of trees build per hour when growing number of samples.}
    \label{figure:synth2.b}
    \vspace{4ex}
  \end{subfigure}
\end{adjustwidth}
\end{figure}

@@

Using a novel parallelization approach enabled by Spark, we extended random forests to cope with extremely high dimensional data (large number of variables). 
We have demonstrated that for data sets with more variable ($p$) than samples ($n$), such as GWAS applications, \cursedforest can perform effective variable selection replicating previous findings and uncovering novel associations in a bone mineral density study. 
We are able to elevate the signal compared to previous approaches using logistic regression due to random forests also modelling interacting features.

@@

To showcase the breadth of application scope, we also demonstrate that \cursedforest's implementation aids in classification problems such as predicting the ethnicity of the 1000 genome project. 
By comparing \cursedforest to other implementations (including those optimised for large data sets) we have demonstrated that runtime can be improved by 80\% allowing analysis in timeframes relevant for point-of-care diagnostics such as for "Patients like mine" scenarios thereby enabling evidence based self learning health care.

@@

% We have demonstrated that using a different parallelization model can extend random forests to the case of an
% extremely large number of variables. We have treated the case of variable selection in a $p >> n$ model where most of
% the variables are uninformative and have demonstrated the utility of the model for large GWAS data sets. By comparing
% this implementation to other implementations (including those optimized for large data sets) we have demonstrated the
% utility of this approach.
